DCMIP2016, Part 1: Models and Equation Sets

### 2017-03-21 DCMIP2016-Part1-Models.tex
- Added edits from James Kent and Vivian Lee / GEM

\affil[1]{University of California, Davis}
\affil[2]{University of Michigan}
\affil[3]{University of South Wales}
\affil[4]{National Center for Atmospheric Research}
\affil[5]{Stony Brook University}
\affil[6]{University of Colorado, Boulder}
\affil[7]{Naval Research Laboratory}
\affil[8]{Colorado State University}
\affil[9]{Geophysical Fluid Dynamics Laboratory}
\affil[10]{Max Planck Institute for Meteorology}
\affil[11]{Deutscher Wetterdienst (DWD)}
\affil[12]{European Centre for Medium-Range Weather Forecasts}
\affil[13]{University of Miami}
\affil[14]{Environment Canada}
\affil[15]{University of Tokyo}
\affil[16]{RIKEN}
\affil[17]{Institut Pierre-Simon Laplace (IPSL)}
\affil[18]{Lawrence Berkeley National Laboratory}

@@

To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment. It was built on previous intercomparison efforts by addressing key outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists. Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling and variable-resolution global modeling. The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.  Modeling groups were then invited to contribute model descriptions and results to the intercomparison effort for publication.

@@

The workshop has advanced our knowledge of (1) the relative behaviors exhibited by atmospheric dynamical cores, (2) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (3) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.  Notably, the use of idealized test cases isolating specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.  A key outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.

@@

\begin{table}[h]
\caption{List of symbols used by DCMIP} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\omega$ & Vertical pressure velocity  \\
$\delta$ & Divergence\\
$\zeta$ & Relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$q$ & Specific humidity \\
$P_{ls}$ & Large-scale precipitation rate \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}[h]
\caption{List of dynamical cores participating in DCMIP which have submitted simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
CSU & Colorado State University Model & Colorado State University \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & Finite Volume Cubed Model & Geophysical Fluid Dynamics Laboratory \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment Canada \\
HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical. DYNAMICO's design uniquely combines:
 \begin{itemize}
\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry,\citep{Dubos2015DYNAMICO10}
\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the litterature,\citep{thuburn2009numerical,ringler2010unified}, and yielding, to a certain extent, an energy-conserving spatial discretization, and
\item a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation). 
\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

The Geophysical Fluid Dynamics Laboratory FV$^3$ model uses a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate. The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms. Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep. Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.

@@

The solver prognoses horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass. The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density. All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.

@@

The GEM model, which is used for operational forecasting at Environment Canada, uses the Yin-Yang grid and horizontal discretization on an Arakawa C grid.  The vertical coordinate is of a log-hydrostatic-pressure type and vertical discretization is based on the Charney-Phillips grid.  A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.  A scalar viscosity is employed for wind components and tracers 
via an application of a scalar Laplacian.  Viscosity operations are applied after the completion of the dynamic time step. {\color{red}[Perhaps expand this text a little bit?]}

@@

The High Order Method Modeling Environment (HOMME) is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.  As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only the shallow-atmosphere are available at present. The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements. These elements are arranged by default in a cubed-sphere structure, but arbitrary quadrilateral grids with conforming edges may also be employed.  The model employs a hybrid terrain-following / pressure coordinate. The fluid equations are discretized using dimension splitting, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions. Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.

@@

The ICON model discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid. The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density. The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves. Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps. The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator. For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion. The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.

@@

Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe.  In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.  The choice of model grid can be motivated by simplicity, as in the case of the latitude-longitude grid, by a desire to maintain a local Cartesian structure, as with the cubed-sphere grid, or to support grid isotropy and homogeneity, as with many of the hexagonal or Voronoi grids that have been employed.  The choice of grid may be further decided by the numerical method -- for instance, most finite element models rely on grids consisting entirely of quadrilaterals.  Inevitably a choice must be made, and the pros and cons of that choice will have repercussions throughout the remainder of the model.  To better understand the options that are available to dynamical core developers, we begin by reviewing many of the model grids that have been employed in global dynamical cores around the world.

@@

The overset Yin-Yang grid \citep{kageama2004yinyang} has two Cartesian grid components (subsets of a latitude-longitude grid) which are geometrically identical (see Figure \ref{fig:gem_yinyang}).  These components are combined to cover a spherical surface with partial overlap along their borders.  The Yin component covers the latitude-longitude region
\begin{equation}\label{eq:gem_yinyang}
(-\frac{\pi}{4}-\delta_{\theta} \leq \theta \leq
\frac{\pi}{4}+\delta_{\theta})  \cap
(-\frac{3\pi}{4}-\delta_{\lambda} \leq \lambda \leq
\frac{3\pi}{4}+\delta_{\lambda}),
\end{equation} where $\delta_{\lambda}, \delta_{\theta}$ are small buffers that are proportional to the respective grid-spacings and are required to enforce a minimum overlap in the overset methodology.  For instance, a common configuration employed by the GEM model for DCMIP fixes 
$ \delta_{\theta}=2 $ degrees and $ \delta_{\lambda}=3 \delta_{\theta}$.  The Yang component covers an analogous area, but is rotated so as to cover the region of the sphere outside of the Yin grid.  This grid is employed by the GEM model, utilizing a pair of regional climate models on the two Cartesian patches.

@@

In this section we describe the fluid equations utilized by nonhydrostatic models.  The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
\begin{align}
\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla.
\end{align}  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
\begin{align} \label{eq:TracerTransport}
\diff{q_i}{t} = 0.
\end{align}

@@

In coordinate-invariant form the prognostic velocity equations may be written in either the Lagrangian or Eulerian frame as
\begin{align} \label{eq:PrognosticVelocity1}
\diff{\vb{u}}{t} = \pdiff{\vb{u}}{t} + \vb{u} \cdot \nabla \vb{u} = - \frac{1}{\rho} \nabla p - 2 \vg{\Omega} \times \vb{u} - \nabla \Phi,
\end{align} where $\vg{\Omega}$ denotes the planetary vorticity vector and $\Phi$ is the geopotential function.  The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.  In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since in an arbitrary coordinate frame this term will give rise to Christoffel symbols associated with derivatives of the vector basis.  Note that it is common to rewrite the pressure gradient force using the relationship
\begin{align}
- \frac{1}{\rho} \nabla p = - \frac{R_d c_p^\ast}{R^\ast} \theta_v \nabla \pi,
\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).  A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.  This gives rise to the 3D vector-invariant form,
\begin{align} \label{eq:PrognosticVelocity2}
\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}. 
\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis that would emerge in (\ref{eq:PrognosticVelocity1}).  In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
\begin{align}
\pdiff{}{t} (\rho \vb{u}) = - \nabla \cdot (\vb{u} \otimes \vb{u} + \mathcal{I} p) - 2 \vg{\Omega} \times (\rho \vb{u}) - \rho \nabla \Phi,
\end{align} where $\vb{u} \otimes \vb{u}$ denotes the outer product and $\mathcal{I}$ is the identity matrix.

@@

{\color{red}[Please assist with fixing notation]}

The vertical coordinate of GEM model (denoted $\zeta$ in the model documentation) is of a log-hydrostatic-pressure type and is defined by
\begin{align}\label{eq:gem_zeta}
 \log \pi_h = s + B(s) \sigma_p, \qquad B(s) =\left( \frac {s-s_T} {(\log p_0) - s_T}\right)^r, \qquad \sigma_p = \log \left( \frac {\pi_h} {p_{0}} \right),
\end{align}
with $s_T= \log \pi_T$ and where $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.

@@

The hydrostatic FV$^3$ model uses a mass-based floating Lagrangian coordinate along with the shallow-atmosphere approximation \citep{L2004MWR}.  Prognostic equations include horizontal velocity in 2D vector-invariant form (\ref{eq:PrognosticUhEquationK2}), pseudo-density (\ref{eq:PrognosticPseudoDensity}), and virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}).  The non-hydrostatic model further incorporates prognostic geopotential (\ref{eq:PrognosticShallowGeopotential}) and vertical momentum (\ref{eq:PrognosticWEquationK2}).

@@

A wide variety of temporal discretizations available among global atmospheric models, particularly to address issues related to the large horizontal-vertical aspect ratio typical in these systems.  In this section we briefly discuss several modern approaches that are presently employed.

>>>>

\affil[1]{University of California, Davis}
\affil[2]{University of Michigan}
\affil[3]{University of South Wales}
\affil[4]{National Center for Atmospheric Research}
\affil[5]{Stony Brook University}
\affil[6]{University of Colorado, Boulder}
\affil[7]{Naval Research Laboratory}
\affil[8]{Colorado State University}
\affil[9]{Geophysical Fluid Dynamics Laboratory}
\affil[10]{Max Planck Institute for Meteorology}
\affil[11]{Deutscher Wetterdienst (DWD)}
\affil[12]{European Centre for Medium-Range Weather Forecasts}
\affil[13]{University of Miami}
\affil[14]{Environment and Climate Change Canada}
\affil[15]{University of Tokyo}
\affil[16]{RIKEN}
\affil[17]{Institut Pierre-Simon Laplace (IPSL)}
\affil[18]{Lawrence Berkeley National Laboratory}

@@

To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment. It was built on earlier DCMIP workshops (held in 2008 and 2012), and previous intercomparison efforts by addressing key outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists. Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling and variable-resolution global modeling. The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.  Modeling groups were then invited to contribute model descriptions and results to the intercomparison effort for publication.

@@

The workshop has advanced our knowledge of (1) the relative behaviors exhibited by atmospheric dynamical cores, (2) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (3) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.  Notably, the use of idealized test cases isolating specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.  A key outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.  The test cases introduced in the 2016 workshop build on the previous DCMIP test case suites \citep{jablonowski2008idealized, ullrich2012dynamical} with tests that now incorporate simplified moist physics.

@@

\begin{table}[h]
\caption{List of symbols used by DCMIP} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$q$ & Specific humidity \\
$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}[h]
\caption{List of dynamical cores participating in DCMIP which have submitted simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
CSU & Colorado State University Model & Colorado State University \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical. DYNAMICO's design uniquely combines:
 \begin{itemize}
\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry,\citep{Dubos2015DYNAMICO10}
\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the litterature,\citep{thuburn2009numerical,ringler2010unified}, and yielding, {\color{red}to a certain extent [what does this mean?]}, an energy-conserving spatial discretization, and
\item a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation). 
\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations. The discretization is done in the horizontal on the equiangular gnomonic cubed-sphere grid \citep{PL2007JCP}, and a flow-following Lagrangian vertical coordinate \citep{lin2004vertically}. The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below). Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid as described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds
that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep. Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.

@@

The prognostic variables are horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass. The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density. All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.

@@

The GEM model \citep{Girard2014}, which is used for operational forecasting at Environment and Climate Change Canada, uses the Yin-Yang grid \citep{kageama2004yinyang} and horizontal discretization on an Arakawa C grid.
The vertical coordinate is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type
and the vertical discretization is based on the Charney-Phillips grid.
A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
It gives rise to an {\it{iterative}} process where each step requires the solution of a
linear system of equations that is reduced to a Helmholtz problem for one composite variable.
For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method on a Yin-Yang grid \citep{Qaddouri2008schwarz}.
The dynamics and physics are time split.
An explicit hyperviscosity is employed for wind components and tracers
via applications of the Laplacian operator.
Viscosity operations are applied after the completion of the physics time step.

@@

The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.  As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only the shallow-atmosphere are available at present. The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements. These elements are arranged by default in a cubed-sphere structure, but arbitrary quadrilateral grids with conforming edges may also be employed.  The model employs a hybrid terrain-following / pressure coordinate. The fluid equations are discretized using dimension splitting, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions. Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.

@@

The ICON model {\color{red}[Reference?]} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid. The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density. The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves. Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps. The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator. For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion. The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.

@@

Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe \citep{staniforth2012horizontal}.  In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.  The choice of model grid can be motivated by simplicity, as in the case of the latitude-longitude grid, by a desire to maintain a local Cartesian structure, as with the cubed-sphere grid, or to support grid isotropy and homogeneity, as with many of the hexagonal or Voronoi grids that have been employed.  The choice of grid may be further decided by the numerical method -- for instance, most finite element models rely on grids consisting entirely of quadrilaterals.  Inevitably a choice must be made, and the pros and cons of that choice will have repercussions throughout the remainder of the model.  To better understand the options that are available to dynamical core developers, we begin by reviewing many of the model grids that have been employed in global dynamical cores around the world.

@@

The overset Yin-Yang grid \citep{kageama2004yinyang} has two Cartesian grid components (subsets of a latitude-longitude grid) which are geometrically identical (see Figure \ref{fig:gem_yinyang}).  These components are combined to cover a spherical surface with partial overlap along their borders.  The Yin component covers the latitude-longitude region
\begin{equation}\label{eq:gem_yinyang}
(-\frac{\pi}{4}-\delta_{\theta} \leq \theta \leq
\frac{\pi}{4}+\delta_{\theta})  \cap
(-\frac{3\pi}{4}-\delta_{\lambda} \leq \lambda \leq
\frac{3\pi}{4}+\delta_{\lambda}),
\end{equation} where $\delta_{\lambda}, \delta_{\theta}$ are small buffers that are proportional to the respective grid-spacings and are required to enforce a minimum overlap in the overset methodology.  For instance, a common configuration employed by the GEM model for DCMIP fixes
$ \delta_{\theta}=2 $ degrees and $ \delta_{\lambda}=3 \delta_{\theta}$.  The Yang component covers an analogous area, but is rotated perpendicularly so as to cover the region of the sphere outside of the Yin grid.  This grid is employed by the GEM model, utilizing a pair of regional climate models on the two Cartesian patches.

@@

In this section we describe the fluid equations utilized by nonhydrostatic models.  The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
\begin{align}
\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla,
\end{align} where $\vb{u}$ denotes the 3D vector velocity.  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
\begin{align} \label{eq:TracerTransport}
\diff{q_i}{t} = 0.
\end{align}

@@

In coordinate-invariant form the prognostic velocity equations may be written in either the Lagrangian or Eulerian frame as
\begin{align} \label{eq:PrognosticVelocity1}
\diff{\vb{u}}{t} = \pdiff{\vb{u}}{t} + \vb{u} \cdot \nabla \vb{u} = - \frac{1}{\rho} \nabla p - 2 \vg{\Omega} \times \vb{u} - \nabla \Phi,
\end{align} where $\vg{\Omega}$ denotes the planetary vorticity vector and $\Phi$ is the geopotential function.  The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.  In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since in an arbitrary coordinate frame this term will give rise to Christoffel symbols associated with derivatives of the vector basis.  Note that it is common to rewrite the pressure gradient force using the relationship
\begin{align}
- \frac{1}{\rho} \nabla p = - \frac{R_d c_p^\ast}{R^\ast} \theta_v \nabla \pi,
\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).  A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.  This gives rise to the 3D vector-invariant form,
\begin{align} \label{eq:PrognosticVelocity2}
\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}. 
\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis  that would arise from the momentum transport term $\vb{u} \cdot \nabla \vb{u}$ in (\ref{eq:PrognosticVelocity1}) (note that these derivatives sometimes appear as Christoffel symbols).  In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
\begin{align}
\pdiff{}{t} (\rho \vb{u}) = - \nabla \cdot (\vb{u} \otimes \vb{u} + \mathcal{I} p) - 2 \vg{\Omega} \times (\rho \vb{u}) - \rho \nabla \Phi,
\end{align} where $\vb{u} \otimes \vb{u}$ denotes the outer product and $\mathcal{I}$ is the identity matrix.

@@

The vertical coordinate of GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.
Taking $s$ (denoted $\pi$ in GEM documentation) as given in (\ref{eq:s_mass-based}), then $\zeta$ is given by the relation
\begin{align}\label{eq:gem_zeta}
\log s = A(\zeta) + B(\zeta) \left[ \log s(z_s)  - \zeta_s \right],
\end{align}
\noindent with $A(\zeta) = \zeta, \, B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r$
and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$
and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.

@@

The hydrostatic FV$^3$ model uses a mass-based floating Lagrangian coordinate along with the shallow-atmosphere approximation \citep{lin2004vertically}.  Prognostic equations include horizontal velocity in 2D vector-invariant form (\ref{eq:PrognosticUhEquationK2}), pseudo-density (\ref{eq:PrognosticPseudoDensity}), and virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}).  The non-hydrostatic model further incorporates prognostic geopotential (\ref{eq:PrognosticShallowGeopotential}) and vertical momentum (\ref{eq:PrognosticWEquationK2}).

@@

A wide variety of temporal discretizations available among global atmospheric models, particularly to address issues related to the large horizontal-vertical aspect ratio typical in these systems.  In this section we briefly discuss several modern approaches that are presently employed.  {\color{red}[reword?]}
